---
title: "Interactive USGS Stream Temperature Data App"
author: "Laura Puckett"
date: "11/29/2021"
output: html_document
---

In this project, I displayed stream temperatures from USGS stations within the Pacific Northwest in an interactive app.

I chose to display these data because a major heatwave struck the region in June 2021. Stream temperatures in many locations grew warmer than 68F - a critical threshold over which salmon experience great stress and are more likely to contract diseases or die. Heat-related salmon mortality was recorded in the region as a result of the 2015 and 2021 heat waves.


- Interactive App [link](https://laurapuckett.shinyapps.io/Pacific_NW_Stream_Temperature_during_2021_Heatwave/)
- Code Repository [link](https://github.com/Laura-Puckett/Stream_Temperature)

My code for this project is shown in a reproducible example below. 

<h2> Part 1. Download Data <h2/>
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
<h3> Load Libraries <h3/>
R version:  `r getRversion()`
```{r Load Libraries, message=FALSE, warning=FALSE}
library(dataRetrieval); library(dplyr); library(shinyr);
library(plotly); library(leaflet)

```
<h3> Define Downloading Functions <h3/>
```{r Define Downloading Functions, eval=FALSE, message=FALSE, warning=FALSE}

# function to download site metadata by state
dl_site_data = function(state){
  print(state)
  sites <- whatNWISdata(stateCd = state, parameterCd="00010", siteType="ST")
  saveRDS(sites, paste('./Rdata/sites', state, sep = '_'))
  return(sites)
}

# function to download stream temperature data by state
dl_temp_data = function(state, startDate, endDate){
  print(paste(state, startDate, endDate))
  streamtemp <- readNWISdata(stateCd = state, service="dv", startDate = startDate,
                             endDate = endDate,parameterCd="00010", siteType="ST")
  streamtemp = streamtemp %>%
    dplyr::select(agency_cd, site_no, dateTime, X_00010_00003, tz_cd)
  saveRDS(streamtemp, paste('./Rdata/streamtemp', startDate, endDate, state, sep = '_'))
  
  return(streamtemp)
}

```
<h3> Download USGS Data <h3/>
```{r Download USGS Data, eval=FALSE}

startDate = '2000-01-01'
endDate = '2021-11-01'
states = c("OR","WA","ID")

# Download Site Metadata
for(state in states){
  # download data or read existing file
  if(file.exists(paste('./Rdata/sites', state, sep = '_'))){
    state_sites = readRDS(paste('./Rdata/sites',  state, sep = '_'))
  }else{
    state_sites = dl_site_data(state)
  }
  # append data from multiple states into single dataframe
  if(state == states[1]){
    site_data = state_sites
  }else{
    site_data = rbind(site_data, state_sites)
    saveRDS(site_data, './Rdata/site_data.rds')
  }
}

# Download Stream Data
for(state in states){
  # download data or read existing file
  if(file.exists(paste('./Rdata/streamtemp', startDate, endDate, state, sep = '_'))){
    state_data = readRDS(paste('./Rdata/streamtemp', startDate, endDate, state, sep = '_'))
  }else{
    state_data = dl_temp_data(state, startDate, endDate)
  }
  # append data from multiple states into single dataframe
  if(state == states[1]){
    temp_data = state_data
  }else{
    temp_data = rbind(temp_data, state_data)
    saveRDS(temp_data, './Rdata/streamtemp_data.rds')
  }
}

```
<h2> Part 2. Organize Data <h2/>

<h3>View Site Metadata<h3/>
```{r View Site Metadata}
site_data = readRDS('./data_prep/Rdata/site_data.rds')
site_data %>% head()

```
<h3>View Stream Temperature Data<h3/>

```{r View Stream Temperature Data}
stream_temp = readRDS('./data_prep/Rdata/streamtemp_data.rds')
stream_temp %>% head()
```
<h3> Reformat Temperature Data <h3/>

```{r}
stream_temp = stream_temp %>% 
  # convert Celsius to Fahrenheit
  rename(tempC = X_00010_00003) %>% 
  mutate(tempF = tempC*(9/5)+32,
         year = lubridate::year(dateTime),
         month = lubridate::month(dateTime),
         month_name = lubridate::month(dateTime, label = T),
         day = lubridate::day(dateTime),
         # reformat dates for plot axes
         month_day = paste0(month_name, day, sep = "-"),
         axis_date_minimal = ifelse(day %in% c(1,15), month_day,""),
         date_all_2021 = dateTime)
# create a dummy column with 2021 as the year for all to easily
# plot multiple years of data on top of each other
lubridate::year(stream_temp$date_all_2021) = 2021
```
<h3> Filtering <h3/>

```{r, eval = F}
# get site_no that have sufficient 2021 data for plotting
sites_with_enough_2021_data = stream_temp %>%
  filter(year == 2021) %>%
  group_by(site_no) %>%
  dplyr::summarize(count = n()) %>%
  filter(count > 300) %>%
  dplyr::select(site_no)

# filter sites dataset
sites_sub = site_data %>%   
  filter(site_no %in% sites_with_enough_2021_data$site_no) %>%
  mutate(huc_cd = as.numeric(huc_cd)) %>%
  filter(end_date>= "2021-08-01" & begin_date <= "2010-04-01") %>%
  filter(count_nu > 1000) %>% # total number of records
  filter(huc_cd >= 17000000 & huc_cd < 18000000) # Colombia River Basin hucs

# filter stream temp dataset
plotData = stream_temp %>% 
  filter(site_no %in% sites_sub$site_no) %>%
  filter(tempF < 140) # remove really high outlier data for one site

# select columns of interest in both datasets
plotData = plotData %>% 
  dplyr::select(site_no, dateTime, tempF, year, month,
                month_day, axis_date_minimal, date_all_2021)

sitesData = sites_sub %>%
  dplyr::select(site_no, station_nm, dec_lat_va, dec_long_va)

# Filter for the data that will be used in each figure on the app
data_for_plot1 = plotData %>%
  filter(dateTime < '2021-09-30' & dateTime > '2021-04-01') %>%
  inner_join(sitesData %>% dplyr::select(site_no, station_nm)) %>%
  unique() %>%
  group_by(site_no)

data_for_plot2 = plotData %>%
  filter(month >= 4 & month <= 9) %>%
  mutate(color_group = "other years",
         color_group = ifelse(year == 2021, 2021, color_group),
         color_group = ifelse(year == 2015, 2015, color_group)) %>%
  inner_join(sitesData %>% dplyr::select(site_no, station_nm)) %>%
  unique() %>%
  group_by(year)

```
<h3> Write Output Files for Shiny App <h3/>
```{r, eval = F}
saveRDS(plotData, './data_prep/Rdata/plotData.rds')
saveRDS(sitesData, './shiny_app/sitesData.rds')
saveRDS(data_for_plot1, './shiny_app/data_for_plot1')
saveRDS(data_for_plot2, './shiny_app/data_for_plot2')
```

<h2> Part 3. The Shiny app <h2/>
<h3> Server Function <h3/>
```{r, eval = F}
server <- function(input, output) {
  # Generate Interactive Map
  output$map <- renderLeaflet(
    map <- leaflet() %>% 
      setView(lng = -117.64, lat = 45.88, zoom = 5.4) %>%
      addProviderTiles(providers$Stamen.TerrainBackground, options = providerTileOptions(opacity = 0.5)) %>%
      addPolygons(data = states_data, fillColor = topo.colors(10, alpha = NULL),
                  stroke = FALSE) %>%
      addMarkers(data = sitesData,
                 ~dec_long_va,
                 ~dec_lat_va, 
                 label = ~htmltools::htmlEscape(station_nm),
                 layerId = ~site_no))
  
  observeEvent(input$map_marker_click, { 
    # save choice from user input in reactive variable
    choice <- input$map_marker_click[1]
  }
  )
  
  output$plot1 <- renderPlotly(
    plot1 <- p_1 %>%
      add_trace(data = data_for_plot1 %>%
                  filter(site_no == ifelse(
                    is.null(input$map_marker_click[1]),
                    14169000,
                    input$map_marker_click[1])),
                type = "scatter",
                x = ~dateTime,
                y = ~tempF,
                mode = 'lines',
                name = "Selected Site",
                line = list(width=2.5, color="black"),
                opacity = 1,
                showlegend = T)
  )
  
  # Generate Plot 2
  output$plot2 <- renderPlotly(
    plot2 <- plot_ly() %>% 
      layout(title = "",
             yaxis=list(title='Stream Temperature [F]'),
             xaxis = list(title = "",
                          type = 'date',
                          tickformat = " %B")) %>%
      add_trace(data = data_for_plot2 %>%
                  filter(year == 2021)  %>%
                  filter(site_no == ifelse(is.null(choice),14169000,choice)),
                name = "2021",
                type = "scatter",
                x = ~date_all_2021,
                y = ~tempF,
                mode = 'lines',
                line = list(width=2.5, color="black",
                            size = 2.5)) %>%
      add_trace(data = data_for_plot2 %>%
                  filter(year == 2015)  %>%
                  filter(site_no == ifelse(is.null(choice),14169000,choice)),
                type = "scatter",
                x = ~date_all_2021,
                name = "2015",s
                y = ~tempF,
                mode = 'lines',
                line = list(width=2.5, color="gray"),
                size = 2.5) %>% 
      add_trace(data = data_for_plot2 %>%
                  filter(color_group == "other years")  %>%
                  filter(site_no == ifelse(is.null(choice),14169000,choice)),
                type = "scatter",
                x = ~date_all_2021,
                y = ~tempF,
                mode = 'lines',
                yaxis="y1",
                name = "other years",
                color =~color_group,
                line=list(color='#1f77b4'),
                opacity=0.4,
                showlegend = T) 
  )
}
```

<h3> User Interface Function <h3/>
```{r, eval = F}
ui <- fluidPage(
  headerPanel('Stream Temperature in the Pacific Northwest'),
  mainPanel(h3("Choose a USGS station from the map. "),
            leafletOutput(outputId = "map"),
            h3(""),
            tabsetPanel(type = "tabs",
                        tabPanel("All Sites, 2021",  plotlyOutput('plot1')),
                        tabPanel("Selected Site, 2000-2021",  plotlyOutput('plot2'))
            )
  )
)
```
<h3> App Setup <h3/>
```{r, eval = F}
heatwave_start = '2021-06-10'
heatwave_end =  '2021-07-20'

data_for_plot1 = readRDS('./shiny_app/data_for_plot1')
data_for_plot2 = readRDS('./shiny_app/data_for_plot2')
sitesData = readRDS('./shiny_app/sitesData.rds')
states_data = maps::map('state', regions=c('oregon', 'washington','idaho'), fill = T, plot = F)

# Initialize plot 1 here (and not in server function) because it only needs to be run once.
p_1 <- plot_ly() %>% 
  layout(# title = "Stream Temperature for All Sites in 2021",
    yaxis=list(title='Stream Temperature [F]'),
    xaxis = list(title = '',
                 type = 'date',
                 tickformat = " %B<br>%Y"),
    shapes = list(list(type = "rect",
                       fillcolor = "red", line = list(color = "red"), opacity = 0.2,
                       y0 = 35, y1 = 85, x0 = heatwave_start, x1 = heatwave_end),
                  list(type = "line", x0 = "2021-04-01", x1 = '2021-09-30',
                       y0 = 68, y1 = 68,
                       line = list(color = "black", dash = "dash")))) %>%
  add_text(x = c("2021-07-01", "2021-05-02"), y = c(38, 71),
           text = c("2021 Heatwave","Dangerous Temperature \n for Salmon"),
           showlegend = F) %>%
  add_trace(data = data_for_plot1,
            type = "scatter",
            x = ~dateTime,
            y = ~tempF,
            mode = 'lines',
            name = ~site_no,
            line=list(width=0.8),
            opacity=.5,
            showlegend = F) 

```
<h3> Run the App <h3/>
```{r, eval = F}
shinyApp(ui = ui, server = server)
```
